Experience Report: Haskell in Computational Biology

### 2012-06-28 mrfy_experience_report.tex
Some cleanup of section 3. NR: please note the missing 'ashore' file in a remark.

A~hidden Markov model carries probabilities on some states and on all
state transitions.
Both the probabilities and the states are determined by the alignment:
\remark{These bullets need proper parallel structure}
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}~$M_j$.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has amino acid~$x$ in
 column~$j$.
This probability is called an \emph{emission} probability.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}~$D_j$.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}~$I_j$.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained amino acid~$x$ by
insertion at column~$j$.
\end{itemize}
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
In~our representation, each state contains a probability or a table of
probabilities, and it is also labelled with one of these labels:
\smallverbatiminput{statelabel}

We~use the ``Plan7'' hidden Markov model, which forbids direct
transitions between insertion 
states and deletion states \cite{Eddy:1998ut}. 
The model gets its name because there are exactly
seven possible transitions
into the states of any column~$j$.
Each transition has its own probability:
\remark{Problem: the subscripts are not so easy to read, and notation
is not quite consistent with \figref{viterbi}, definition of $a'$ at
bottom right.}
\begin{itemize} 
\item
A~transition into a match state 
is more likely when column~$j$ is a consensus column.
Depending on the predecessor state, 
the probability of such a transition is 
$\txprobj M M$, $\txprobj I M$, or~$\txprobj D M$.
\item
A~transition into a deletion state 
is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobj M D$~or~$\txprobj D D$.
\item
A~transition into an insertion state 
is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobjj M I$~or~$\txprobjj I I$.
\end{itemize}

@@

\def\goo{18pt}
\def\gum{14pt}
\[
\begin{array}{l@{}c@{}l}
V_{j}^{M}(i) &{}={}& \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j-1}^{M}(i - 1)} {\log a_{M_{j-1}M_{j}}}\\
\vsum{V_{j-1}^{I}(i - 1)} {\log a_{I_{j-1}M_{j}}}\\
\vsum{V_{j-1}^{D}(i - 1)} {\log a_{D_{j-1}M_{j}}}\\
\end{array} \right.\\[\goo]
V_{j}^{I}(i) &=& \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j}^{M}(i - 1)} {\log a_{M_{j}I_{j}}}\\
\vsum{V_{j}^{I}(i - 1)} {\log a_{I_{j}I_{j}}}\\
\end{array} \right.\\[\gum]
V_{j}^{D}(i) &=& \max \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j-1}^{M}(i)} {\log a_{M_{j-1}D_{j}}}\\
\vsum{V_{j-1}^{D}(i)} {\log a_{D_{j-1}D_{j}}}\\
\end{array} \right.\\
\end{array}
\qquad
\begin{array}{l@{}c@{}l}
V_{j}^{\prime M}(i) &{}={}& e^{\prime}_{M_{j}}(x_{i}) + \min \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j-1}^{\prime M}(i - 1)} {a^{\prime}_{M_{j-1}M_{j}}}\\
\vsum{V_{j-1}^{\prime I}(i - 1)} {a^{\prime}_{I_{j-1}M_{j}}}\\
\vsum{V_{j-1}^{\prime D}(i - 1)} {a^{\prime}_{D_{j-1}M_{j}}}\\
\end{array} \right.\\[\goo]
V_{j}^{\prime I}(i) &=& e^{\prime}_{I_{j}}(x_{i}) + \min \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j}^{\prime M}(i - 1)} {a^{\prime}_{M_{j}I_{j}}}\\
\vsum{V_{j}^{\prime I}(i - 1)} {a^{\prime}_{I_{j}I_{j}}}\\
\end{array} \right.\\[\gum]
V_{j}^{\prime D}(i) &=& \min \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j-1}^{\prime M}(i)} {a^{\prime}_{M_{j-1}D_{j}}}\\
\vsum{V_{j-1}^{\prime D}(i)} {a^{\prime}_{D_{j-1}D_{j}}}\\
\end{array} \right.\\[\gum]
\multicolumn3{l}{%
a'_{s} = - \log a_{s} 
\qquad
e^{\prime}_{s,x} = - \log\frac{e_{s,x}}{q_{x}}
\qquad
V_j^{\prime M}(i) = - V_j^{M}(i)
}
\\
\end{array}
\]

@@

In a hidden Markov model, each transition is actually labeled with a
score, not a probability.
And we include the labels of the states that are
connected:\remark{\emph{Why} are the labels there?
And since they are apparently never used, why do they have strictness annotations???}
\begin{smallverbatim}
data TProb =   --- THIS DEFN IS A DAMN LIE
     TProb { logProbability :: Score
           , fromState :: StateLabel
           , toState :: StateLabel
           }
\end{smallverbatim}
In~a Plan7 hidden Markov model, as discussed above, 
there are seven transitions into each node:
\begin{smallverbatim}
data TProbs =  -- MORE DAMN LIES
     TProbs { m_m :: TProb, m_i :: TProb, m_d :: TProb
            , i_m :: TProb, i_i :: TProb
            , d_m :: TProb, d_d :: TProb }
\end{smallverbatim}
In~our code, we select from this 7-tuple using the \texttt{edge}
function:\remark{Broken!  This function no longer appears in our examples}
\smallverbatiminput{edge}
In MRFy, the first two arguments to \texttt{edge} are known at compile
time, and we expect \texttt{edge} to be inlined.

@@

Another of us was curious to try coding Viterbi's equations
directly as recursive functions.
Like the classic recursive Fibonacci function, Viterbi's functions,
when implemented \naive ly,
perform extraordinarily badly.
But just like the Fibonacci function, Viterbi's functions can be
\emph{memoized}.
Our code implements a transformed version of Viterbi's equations, 
which operates on scores instead of on probabilities.
This version is
shown on the right-hand side of \figref{viterbi}.
THIS PART IS BROKEN.  THE A ARRAY NEEDS TWO SUBSCRIPTS, AND THE E
NOTATION IS INCONSISTENT (HAS MAGICALLY CHANGED FROM BRACKETS TO SUBSCRIPTS).
The transformed equations minimize the
score (i.e., the negated log probability) for each combination of
node~$j$, amino 
acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.

@@

We~also made \texttt{Scored} an instance of~\texttt{Ord} and of
\texttt{Functor}, so
\begin{smallverbatim}
fmap :: (a -> b) -> (Scored a -> Scored b)
\end{smallverbatim}
applies a function to the thing inside the container, without touching
the score.\remark{Changes in the code appear to have
made \texttt{fmap} irrelevant}
And containers are ordered by score alone, so applying
\texttt{minimum} to a list of scored things chooses the thing with the
smallest (and therefore best) score.

The \texttt{Scored} abstraction makes it easy to implement
Viterbi's equations.
For example, to compute $V_j^{\prime M}(i)$ using the equation at the top
right of \figref{viterbi}, we call 
\mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}, 
which chooses the best (\texttt{minimum}) solution
from among the three predecessor states:
\smallverbatiminput{vfix}
In this code, EXPLAIN EXTEND, EXPLAIN ESCORE, EXPLAIN
ASCORE. 
we~use~\texttt{/+/} to add transition and emission scores,
and we use \texttt{fmap} to add \texttt{Mat} to the list of states
encountered up to this point:\remark{BROKEN! \texttt{fmap} is gone}

>>>>

A~hidden Markov model carries probabilities on some states and on all
state transitions.
Both the probabilities and the states are determined by the alignment:
\remark{These bullets need proper parallel structure}
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}~$M_j$.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has amino acid~$x$ in
 column~$j$.
This probability is called an \emph{emission} probability.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}~$I_j$.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained amino acid~$x$ by
insertion at column~$j$.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}~$D_j$.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
\end{itemize}
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
In~our representation, each state contains a probability or a table of
probabilities, and it is also labelled with one of these labels:
\smallverbatiminput{statelabel}

We~use the ``Plan7'' hidden Markov model, which forbids direct
transitions between insertion 
states and deletion states \cite{Eddy:1998ut}. 
The model gets its name because there are exactly
seven possible transitions
into the states of any column~$j$.
Each transition has its own probability:
\remark{Norman: how do you propose to make the subscripts more readable?}
\begin{itemize} 
\item
A~transition into a match state 
is more likely when column~$j$ is a consensus column.
Depending on the predecessor state, 
the probability of such a transition is 
$\txprobj M M$, $\txprobj I M$, or~$\txprobj D M$.
\item
A~transition into a deletion state 
is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobj M D$~or~$\txprobj D D$.
\item
A~transition into an insertion state 
is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobjj M I$~or~$\txprobjj I I$.
\end{itemize}

@@

\def\goo{18pt}
\def\gum{14pt}
\[
\begin{array}{l@{}c@{}l}
V_{j}^{M}(i) &{}={}& \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j-1}^{M}(i - 1)} {\log a_{M_{j-1}M_{j}}}\\
\vsum{V_{j-1}^{I}(i - 1)} {\log a_{I_{j-1}M_{j}}}\\
\vsum{V_{j-1}^{D}(i - 1)} {\log a_{D_{j-1}M_{j}}}\\
\end{array} \right.\\[\goo]
V_{j}^{I}(i) &=& \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j}^{M}(i - 1)} {\log a_{M_{j}I_{j}}}\\
\vsum{V_{j}^{I}(i - 1)} {\log a_{I_{j}I_{j}}}\\
\end{array} \right.\\[\gum]
V_{j}^{D}(i) &=& \max \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j-1}^{M}(i)} {\log a_{M_{j-1}D_{j}}}\\
\vsum{V_{j-1}^{D}(i)} {\log a_{D_{j-1}D_{j}}}\\
\end{array} \right.\\
\end{array}
\qquad
\begin{array}{l@{}c@{}l}
V_{j}^{\prime M}(i) &{}={}& e^{\prime}_{M_{j}}(x_{i}) + \min \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j-1}^{\prime M}(i - 1)} {a^{\prime}_{M_{j-1}M_{j}}}\\
\vsum{V_{j-1}^{\prime I}(i - 1)} {a^{\prime}_{I_{j-1}M_{j}}}\\
\vsum{V_{j-1}^{\prime D}(i - 1)} {a^{\prime}_{D_{j-1}M_{j}}}\\
\end{array} \right.\\[\goo]
V_{j}^{\prime I}(i) &=& e^{\prime}_{I_{j}}(x_{i}) + \min \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j}^{\prime M}(i - 1)} {a^{\prime}_{M_{j}I_{j}}}\\
\vsum{V_{j}^{\prime I}(i - 1)} {a^{\prime}_{I_{j}I_{j}}}\\
\end{array} \right.\\[\gum]
V_{j}^{\prime D}(i) &=& \min \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j-1}^{\prime M}(i)} {a^{\prime}_{M_{j-1}D_{j}}}\\
\vsum{V_{j-1}^{\prime D}(i)} {a^{\prime}_{D_{j-1}D_{j}}}\\
\end{array} \right.\\[\gum]
\multicolumn3{l}{%
a'_{s_{j}s_{k}} = - \log a_{s_{j}s_{k}} 
\qquad
e^{\prime}_{s}(x) = - \log\frac{e_{s}(x)}{q_{x}}
\qquad
V_j^{\prime M}(i) = - V_j^{M}(i)
}
\\
\end{array}
\]

@@

In a hidden Markov model, each transition is actually labeled with a
score, not a probability.
And we include the labels of the states that are
connected:\remark{\emph{Why} are the labels there?
And since they are apparently never used, why do they have strictness annotations???}
\begin{smallverbatim}
data TProb =
     TProb { logProbability :: Score
           , fromState :: StateLabel
           , toState :: StateLabel
           }
\end{smallverbatim}
In~a Plan7 hidden Markov model, as discussed above, 
there are seven transitions into each node:
\begin{smallverbatim}
data TProbs =
     TProbs { m_m :: TProb, m_i :: TProb, m_d :: TProb
            , i_m :: TProb, i_i :: TProb
            , d_m :: TProb, d_d :: TProb }
\end{smallverbatim}
In~our code, we select from this 7-tuple using the \texttt{ascore}
function:\remark{Norman: please put ascore into a file}
\smallverbatiminput{ascore}
In MRFy, the first two arguments to \texttt{ascore} are known at compile
time, and we expect \texttt{ascore} to be inlined.

@@

Another of us was curious to try coding Viterbi's equations
directly as recursive functions.
Like the classic recursive Fibonacci function, Viterbi's functions,
when implemented \naive ly,
perform extraordinarily badly.
But just like the Fibonacci function, Viterbi's functions can be
\emph{memoized}.
Our code implements a transformed version of Viterbi's equations, 
which operates on scores instead of on probabilities.
This version is
shown on the right-hand side of \figref{viterbi}.
The transformed equations minimize the
score (i.e., the negated log probability) for each combination of
node~$j$, amino 
acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.

@@

We~also made \texttt{Scored} an instance of~\texttt{Ord}.
Containers are ordered by score alone, so applying
\texttt{minimum} to a list of scored things chooses the thing with the
smallest (and therefore best) score.

The \texttt{Scored} abstraction makes it easy to implement
Viterbi's equations.
For example, to compute $V_j^{\prime M}(i)$ using the equation at the top
right of \figref{viterbi}, we call 
\mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}, 
which chooses the best (\texttt{minimum}) solution
from among the three predecessor states:
\smallverbatiminput{vfix}

In this code, we look up emission scores for a given node and amino
acid using \texttt{eScore},
we~use~\texttt{/+/} to add transition and emission scores,
and we use \texttt{extend} to add \texttt{Mat} to the list of states
encountered up to this point.
